---
title: "Getting started with ShuttleboxR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ShuttleboxR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  warning = FALSE,
  message = FALSE
)
```

## What ShuttleboxR does

The aim of ShuttleboxR is to take the raw shuttle-box recordings to a checked and
interpretable project dataset. The workflow has three stages:

1. **Import and organise** the data.
2. **Calculate shuttle-box metrics** consistently.
3. **Inspect and troubleshoot** individual trials and the complete project.

The package works at two linked levels:

- **one fish**, where the full time series can be inspected; and
- **the complete project**, where all fish can be compared and unusual trials
  flagged for closer review.

A project-level flag of an individual trial is not an automatic reason for exclusion. The intended
workflow is **flag → inspect the raw trial → decide and document**.

## Installation

ShuttleboxR can be installed directly from GitHub. First install the
`remotes` package if it is not already available:

```{r installation, eval=FALSE}
install.packages("remotes")
```

Then install ShuttleboxR:

```{r install-shuttleboxr, eval=FALSE}
remotes::install_github(
  "PBRiesenkamp/ShuttleboxR",
  upgrade = "never"
)
```

Load the package before beginning an analysis:

```{r load-package}
library(ShuttleboxR)
```

To install the package together with its vignette, use:

```{r install-with-vignette, eval=FALSE}
remotes::install_github(
  "PBRiesenkamp/ShuttleboxR",
  upgrade = "never",
  build_vignettes = TRUE
)
```

The vignette can then be opened in R using:

```{r open-vignette, eval=FALSE}
vignette("ShuttleboxR", package = "ShuttleboxR")
```

To update an existing installation to the latest version, reinstall the
package with `force = TRUE`:

```{r update-package, eval=FALSE}
remotes::install_github(
  "PBRiesenkamp/ShuttleboxR",
  force = TRUE,
  upgrade = "never"
)
```

## Choose your starting point

| Starting data | Function | Result |
|---|---|---|
| One raw ShuttleSoft file | `read_shuttlesoft()` | `fish`: one time-series trial |
| A folder of raw files | `read_shuttlesoft_project()` | `all_fish`: a list of trials |
| An existing one-row-per-fish summary CSV | `read_project_database()` | `project_data`: a project summary table |

### One raw file

```{r starting-point-single, eval=FALSE}
fish <- read_shuttlesoft(file.choose())
```

### A folder of raw files

```{r starting-point-folder, eval=FALSE}
all_fish <- read_shuttlesoft_project(directory = choose.dir())
project_data <- calc_project_results(all_fish)
```

The Windows folder-selection window displays folders rather than the `.txt` and
`.csv` files inside them. Select the folder containing all raw trials.

### An existing project summary

```{r starting-point-summary, eval=FALSE}
project_data <- read_project_database(file.choose())
```

Do not use `read_shuttlesoft()` for an existing one-row-per-fish table. It
expects raw time-series observations.

# Branch 1: analyse and inspect one fish

## Import the recording

```{r choose-file, eval=FALSE}
fish <- read_shuttlesoft(file.choose())
```

When the recording contains an acclimation period, provide the time at which the
dynamic trial started:

```{r trial-start, eval=FALSE}
fish <- read_shuttlesoft(file.choose(), trial_start = "13:30:00")
```

This vignette uses the example recording included with the package:

```{r import-example}
example_file <- system.file(
  "extdata",
  "Fish_70_13_2_rev_B.txt",
  package = "ShuttleboxR"
)
fish <- read_shuttlesoft(example_file)
inspect(fish)
```

`read_shuttlesoft()` prepares the file automatically. It creates elapsed time,
identifies shuttle events and retains the `core_T` values already produced by
ShuttleSoft.

## View gravitation before calculating thermal metrics

The beginning of a dynamic trial may describe the route from the starting
temperature to the fish's settled thermal region. Including this transition period
can pull Tpref towards the starting conditions and bias Tbreadth, avoidance
percentiles, core-temperature variation, and apparent time near limits.

View the fitted transition first:

```{r gravitation-plot, fig.height=5.2}
plot_T_segmented(fish)
```

`plot_T_segmented()` estimates gravitation automatically, shades the transition
and marks the breakpoint. The same duration can be reported numerically with:

```{r gravitation-value}
calc_gravitation(fish)
```

The user does not need to save this value or pass it to every later function.
When `exclude_gravitation = TRUE`in the relevant functions, that function estimates 
gravitation under the hood and applies the correct cutoff for each specific trial.

Inspect the segmented plot before relying on the breakpoint. Continued drift,
an implausible breakpoint, little data after the breakpoint, temperature-control
failure, or prolonged contact with the programmed limits all deserve closer
attention.

## Calculate settled thermal metrics

```{r basic-thermal-metrics}
calc_Tpref(fish, exclude_gravitation = TRUE)
calc_Tavoid(fish, exclude_gravitation = TRUE)
calc_Tbreadth(fish, exclude_gravitation = TRUE)
calc_Tpercentile_range(fish, exclude_gravitation = TRUE)
```

The same simple structure applies to other thermal summaries:

```{r additional-thermal-metrics}
calc_extremes(fish, exclude_gravitation = TRUE)
calc_coreT_variance(
  fish,
  variance_type = "std_deviation",
  exclude_gravitation = TRUE
)
```

When `trial_start` was supplied and only the dynamic period should be analysed,
add `exclude_acclimation = TRUE`:

```{r exclude-acclimation-example, eval=FALSE}
calc_Tpref(
  fish,
  exclude_acclimation = TRUE,
  exclude_gravitation = TRUE
)
```

Thermal metrics calculated after gravitation describe **settled thermal
behaviour** rather than the time and route taken to reach it.

Distance, shuttling and chamber occupancy can describe either the complete
selected period or only settled thermoregulation:

```{r activity-window-example, eval=FALSE}
# Complete selected period
calc_shuttles(fish)
calc_tot_distance(fish)

# Settled thermoregulation only
calc_shuttles(fish, exclude_gravitation = TRUE)
calc_tot_distance(fish, exclude_gravitation = TRUE)
```

The optional `gravitation_time` argument remains available for advanced use when
a manually checked value must be reused exactly. It is not part of the normal
workflow.

## What Tpref and the thermal-spread metrics mean

- `Tpref` describes the centre of the selected temperature distribution. The
  default is the median `core_T`.
- `Tavoid_lower` and `Tavoid_upper` are the lower and upper percentile
  boundaries. Their difference is `Tpref_range`.
- `Tpercentile_range` is the difference between two selected percentiles.
  The defaults are the 25th and 75th percentiles, giving the interquartile
  range: the width of the central 50% of observations.
- `Tbreadth` describes how far apart the temperatures experienced by the fish
  were across the complete distribution.


### Percentile-based thermal range

The percentile-based range is calculated as:

\[
Tpercentile\_range = Q_{upper} - Q_{lower}
\]

With the defaults, this is `Q75 - Q25`. A value of 1.5 °C means that the middle
half of the fish's core-temperature observations spans 1.5 °C. It does not
mean that the interval is centred on Tpref.

```{r percentile-range-example}
calc_Tpercentile_range(fish, exclude_gravitation = TRUE)
```

The central interval can be widened or narrowed by changing the percentiles:

```{r percentile-range-custom, eval=FALSE}
calc_Tpercentile_range(
  fish,
  percentiles = c(0.10, 0.90),
  exclude_gravitation = TRUE
)
```

This measure is less influenced by the tails than `Tbreadth`, because it uses
only the two chosen percentile boundaries. It is also conceptually distinct
from `Tpref_range`, which is derived from the avoidance percentiles specified in
`calc_Tavoid()`.

### Tbreadth explained

Imagine choosing two moments from the analysed part of the trial and comparing
the fish's core temperature at those moments. Repeat this for every possible
pair and take the average difference. That average is `Tbreadth`.

- A fish remaining near one temperature has a Tbreadth close to 0 °C.
- Equal time at 20 °C and 21 °C gives a Tbreadth of 0.5 °C.
- Equal time at 10 °C and 20 °C gives a Tbreadth of 5 °C.

For core-temperature observations \(T_1, T_2, \ldots, T_n\):

\[
Tbreadth = \frac{1}{n^2}
\sum_{i=1}^{n}\sum_{j=1}^{n}|T_i-T_j|
\]

`Tbreadth` is not centred on Tpref and does not define a lower and upper
interval. It also cannot show histogram shape by itself. Two fish can have the
same Tbreadth even when one distribution has one broad peak and the other has
two separate peaks.

View the value together with the histogram:

```{r core-temperature-histogram, fig.height=4.6}
plot_coreT_histogram(fish, exclude_gravitation = TRUE)
```

The histogram's `bin_size` changes only the drawing. It does not change the
Tbreadth calculation.

## Inspect whether the trial is interpretable

Inspection should answer two questions:

1. Did the system and tracking work as intended?
2. Did the fish show behaviour that can reasonably be interpreted as
   thermoregulation?

### Chamber temperatures

```{r plot-gradient, fig.height=4.4}
plot_T_gradient(fish)
```

Look for a stable warm-cold difference, plausible heating and cooling rates and
no abrupt interruptions or impossible values.

### Tracking and spatial behaviour

```{r tracking-plot, fig.height=4.4}
plot_tracking(fish)
```

```{r heatmap-plot, fig.height=5.2}
plot_heatmap(fish)
```

Poor tracking can inflate or suppress distance and create false shuttles. A
heatmap concentrated in one small area can indicate inactivity, tracking lock 
on an object, or a fish that did not engage with the thermal gradient.

### Numerical summaries of movement and data quality

Several useful trial-level summaries can be obtained directly:

```{r single-trial-summary-values}
calc_shuttles(fish)
calc_tot_distance(fish)
calc_occupancy(fish)
calc_track_accuracy(fish)
```

`calc_shuttles()` counts chamber changes, `calc_tot_distance()` reports total
movement, `calc_occupancy()` compares use of the warm and cold chambers, and
`calc_track_accuracy()` reports the proportion of observations with a valid
track. Together, these values provide a quick check before interpreting more
complex thermal metrics.

### Distance moved through time

```{r distance-over-time, fig.height=4.6}
plot_distance(fish)
```

The cumulative-distance curve shows when movement occurred. A steep section
indicates rapid accumulation of distance, whereas a plateau indicates little or
no movement. Sudden unrealistic jumps can suggest tracking errors.

### Activity in time blocks

```{r interval-activity, fig.height=4.6, results='hide'}
plot_interval(
  fish,
  column = "velocity",
  interval_minutes = 30
)
```

`plot_interval()` divides the trial into user-selected time blocks and plots the
mean of any suitable column. Here it shows mean velocity in 30-minute blocks,
making prolonged changes in activity easier to see than in the second-by-second
record. Other columns can be selected in the same way.

### Movement speed across core temperatures

```{r speed-vs-core-temperature, fig.height=4.8}
plot_speed_coreT(fish)
```

This plot is useful for asking whether activity changes across the temperatures
experienced by the fish. It should be treated as an exploratory relationship,
particularly when observations are strongly ordered through time.

### Additional exploratory views

The generic histogram function can display the distribution of any suitable
trial variable, and the movement record can also be animated:

```{r additional-exploratory-plots, eval=FALSE}
plot_histogram(fish, column = "velocity", binwidth = 0.5)
animate_movements(fish)
```

## Single-trial function finder

| Question | Useful function |
|---|---|
| Did the temperature system behave correctly? | `plot_T_gradient()` |
| Where is the gravitation breakpoint? | `plot_T_segmented()`, `calc_gravitation()` |
| What is the centre and shape of temperature use? | `plot_coreT_histogram()`, `calc_Tpref()`, `calc_Tbreadth()`, `calc_Tpercentile_range()` |
| When did movement occur? | `plot_distance()`, `plot_interval()` |
| Did activity vary with temperature? | `plot_speed_coreT()` |
| Was tracking reliable? | `plot_tracking()`, `calc_track_accuracy()` |
| Where did the fish spend its time? | `plot_heatmap()`, `calc_occupancy()` |
| How much did the fish move and shuttle? | `calc_tot_distance()`, `calc_shuttles()` |

# Branch 2: screen the complete project

## Create or import `project_data`

### Route A: start from raw trial files

```{r project-import, eval=FALSE}
all_fish <- read_shuttlesoft_project(directory = choose.dir())

project_data <- calc_project_results(
  all_fish,
  exclude_gravitation_thermal = TRUE
)
```

`calc_project_results()` estimates gravitation once per fish and reuses it for
all thermal metrics, including `Tpercentile_range`. Activity metrics retain
the complete selected period unless
`exclude_gravitation_activity = TRUE` is also specified.

When the raw files have different dynamic-period start times, pass an optional
metadata table to `read_shuttlesoft_project()`. This is an additional import
option, not a required part of the standard workflow.

### Route B: start from an existing project-summary CSV

```{r existing-project-database, eval=FALSE}
project_data <- read_project_database(file.choose())
```

An existing summary file can only provide metrics already stored in it. For
example, neither `Tbreadth` nor `Tpercentile_range` can be reconstructed from
Tpref and avoidance temperatures. They must already be present in the summary
table or be recalculated from the raw trials.

The package includes an example project database:

```{r import-project-database}
project_file <- system.file(
  "extdata",
  "project_database_example.csv",
  package = "ShuttleboxR"
)
project_data <- read_project_database(project_file)
```

## Begin with single-variable distributions

```{r project-histograms, fig.height=9, fig.width=8, results='hide'}
plot_histograms(project_data)
```

These histograms identify fish that are unusual for individual metrics. A
univariate flag does not reveal whether the value is technically invalid or
biologically meaningful.

## Inspect upper and lower limit exposure separately

```{r upper-lower-extremes, fig.height=5.4}
plot_upper_vs_lower_extremes(project_data)
```

The default dashed lines mark 10% of analysed observations near each limit. This
absolute threshold remains interpretable when most fish have zero exposure.
The threshold can be changed directly:

```{r upper-lower-thresholds, eval=FALSE}
plot_upper_vs_lower_extremes(
  project_data,
  lower_limit_threshold = 5,
  upper_limit_threshold = 5
)
```

A fish above an upper, lower or both limit thresholds should be returned to
`plot_T_gradient()`, `plot_T_segmented()` and `plot_coreT_histogram()`.

## Use bivariate plots to identify different warning patterns

### Distance versus shuttling

```{r distance-shuttle-review, fig.height=5.8, fig.width=7.5}
plot_distance_vs_shuttles(project_data, highlight_cases = TRUE)
```

Each point represents one fish. This plot is a quick way to find individuals
whose total movement and number of chamber changes form an unusual combination.

Useful patterns include:

- **low movement + low shuttling:** inspect inactivity and tracking;
- **low movement + high shuttling:** inspect doorway crossings and false
  shuttles;
- **high movement + low shuttling:** active movement within one chamber; and
- **high movement + high shuttling:** active thermoregulation or possible
  agitation.

### Limit exposure versus activity

```{r limits-distance-review, fig.height=5.8}
plot_limits_vs_distance(project_data, highlight_cases = TRUE)
```

```{r limits-shuttles-review, fig.height=5.8}
plot_limits_vs_shuttles(project_data, highlight_cases = TRUE)
```

Exposure above the threshold with little activity suggests that the fish may
have remained near a limit without responding. High exposure with high activity
may indicate active searching within an unsuitable programmed range.

The plot is all most users need. To obtain the table of highlighted fish, add
`return_cases = TRUE`:

```{r retrieve-cases, eval=FALSE}
review <- plot_distance_vs_shuttles(
  project_data,
  highlight_cases = TRUE,
  return_cases = TRUE
)
review$cases
```

## View all pairwise relationships

```{r project-bivariate-matrix, fig.height=8, fig.width=8}
correlation_matrix(
  project_data,
  columns = c(
    "Tpref", "Tpref_range", "grav_time",
    "tot_distance", "nr_shuttles", "t_near_limits"
  )
)
```

Variable names appear on the diagonal. The lower panels show pairwise
scatterplots and smooth trends; the upper panels show Pearson correlations. A
point separated from the main cloud in one panel may represent an unusual
combination of otherwise plausible measurements.

The matrix is also useful for identifying strongly correlated variables before
PCA.

When available, `Tpercentile_range` can be added to the selected columns. Check
its correlation with `Tpref_range` and `Tbreadth` before including several
thermal-spread metrics in the same PCA.

## Use PCA to screen multivariate behaviour

PCA results depend on the variables included. Supply them explicitly so that a
new numeric column cannot silently change the analysis.

```{r project-pca, fig.height=5.8, fig.width=7.5}
project_pca <- pca(
  project_data,
  variables = c(
    "Tpref", "Tpref_range", "grav_time",
    "tot_distance", "nr_shuttles",
    "t_near_max", "t_near_min"
  ),
  print_labels = FALSE
)

project_pca$plots$screeplot
```

The scree plot shows how much project variation is represented by each
principal component. PC1-PC2 is only a two-dimensional view; later components
can contain additional unusual structure.

```{r pca-biplot, fig.height=6.4, fig.width=8}
project_pca$plots$biplot
```

Read the biplot as follows:

- points close together have similar combinations of metrics;
- an arrow points towards increasing values of that metric;
- longer arrows are better represented in the visible PC1-PC2 plane;
- arrows in similar directions indicate positive association; and
- a fish far in an arrow's direction is likely to have a relatively high value
  for that metric.

The default screen is quite conservative. A fish is highlighted only
when detected by both Mahalanobis distance and DBSCAN. The detailed output is
available directly:

```{r pca-details}
project_pca$outlier_details
```

Sensitivity can be changed with `mahalanobis_th`, `dbscan_th`,
`dbscan_minPts` and `flag_rule`. For example, `flag_rule = "either"` is less
conservative than the default `"both"`. These settings change which fish are
highlighted; they do not prove that any fish is invalid.

## Return flagged fish to the raw trial

```{r review-loop, eval=FALSE}
fish_to_check <- all_fish[["Fish_17.txt"]]

plot_T_gradient(fish_to_check)
plot_tracking(fish_to_check)
plot_T_segmented(fish_to_check)
plot_coreT_histogram(fish_to_check, exclude_gravitation = TRUE)
plot_heatmap(fish_to_check)
plot_distance(fish_to_check)

calc_Tpref(fish_to_check, exclude_gravitation = TRUE)
calc_Tavoid(fish_to_check, exclude_gravitation = TRUE)
calc_Tbreadth(fish_to_check, exclude_gravitation = TRUE)
calc_Tpercentile_range(fish_to_check, exclude_gravitation = TRUE)
calc_extremes(fish_to_check, exclude_gravitation = TRUE)
```

No manual gravitation object is required. Each post-gravitation function uses
the same simplified argument and performs the necessary calculation internally.

A suggested decision process is:

1. **Flag** an unusual metric or combination of metrics.
2. **Inspect** the original temperature, tracking and movement records.
3. **Identify a reason**, such as a power interruption, tracking failure,
   prolonged immobility, unsuitable limits or behaviour inconsistent with the
   intended assay.
4. **Compare** the pattern with the rest of the project and the ecology of the
   species.
5. **Document** whether the fish is retained, excluded or included in a
   sensitivity analysis.

A fish should not be excluded merely because it is statistically unusual.
Unusual values can be interesting biological variation!

## Getting help

```{r help-pages, eval=FALSE}
?calc_Tbreadth
?calc_Tpercentile_range
?plot_distance_vs_shuttles
?pca
help(package = "ShuttleboxR")
```
