---
title: "Getting started with ShuttleboxR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ShuttleboxR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

## What ShuttleboxR does

ShuttleboxR provides a single workflow for moving from a raw ShuttleSoft file
to calculated thermal and behavioural metrics. It can:

- import ShuttleSoft `.txt` files and comma-separated `.csv` exports;
- label acclimation and trial periods;
- retain the `core_T` values already generated by ShuttleSoft;
- calculate thermal preference, avoidance temperatures, thermal breadth,
  shuttling, chamber occupancy, movement, and tracking quality;
- create plots for individual trials; and
- summarise all trials in a project.

```{r load-package}
library(ShuttleboxR)
```

## Import one file

The simplest way to import your own file is:

```{r choose-file, eval=FALSE}
fish <- read_shuttlesoft(file.choose())
```

`file.choose()` opens an ordinary file-selection window. The returned data are
prepared automatically, so calculation and plotting functions can be used
immediately.

This vignette uses a shortened ShuttleSoft recording included with the package:

```{r import-example}
example_file <- system.file(
  "extdata",
  "Fish_8_13_3_example.csv",
  package = "ShuttleboxR"
)

fish <- read_shuttlesoft(example_file)
```

Check the size and first few column names:

```{r inspect-structure}
dim(fish)
head(names(fish), 12)
```

You can also run:

```{r inspect-file}
inspect(fish)
```

## Mark acclimation and trial periods

When `trial_start` is omitted, the first observation is treated as the start of
the trial and the whole recording is labelled `"trial"`.

When you know the clock time at which the experimental trial began, provide it
during import:

```{r trial-start, eval=FALSE}
fish <- read_shuttlesoft(
  file.choose(),
  trial_start = "13:30:00"
)
```

Rows before that time will be labelled `"acclimation"`. You can then exclude
them consistently from calculations and plots:

```{r exclude-acclimation, eval=FALSE}
calc_Tpref(fish, exclude_acclimation = TRUE)
calc_Tbreadth(fish, exclude_acclimation = TRUE)
plot_coreT_histogram(fish, exclude_acclimation = TRUE)
```

## Calculate thermal metrics

### Preferred temperature

The default estimate of preferred temperature is the median `core_T`:

```{r tpref}
Tpref <- calc_Tpref(
  fish,
  method = "median",
  print_results = FALSE
)
Tpref
```

The mean and unbinned mode are also available:

```{r tpref-methods}
calc_Tpref(fish, method = "mean", print_results = FALSE)
calc_Tpref(fish, method = "mode", print_results = FALSE)
```

### Avoidance temperatures

`calc_Tavoid()` returns lower and upper percentiles of the observed `core_T`
distribution. The defaults are the 5th and 95th percentiles:

```{r tavoid}
Tavoid <- calc_Tavoid(
  fish,
  percentiles = c(0.05, 0.95),
  print_results = FALSE
)
Tavoid
```

The simple central range is therefore:

```{r tavoid-range}
Tavoid[2] - Tavoid[1]
```

### Effective selected thermal breadth

`calc_Tbreadth()` uses the complete frequency distribution rather than only its
upper and lower limits. Temperatures are divided into fixed-width bins and the
breadth is calculated as:

\[
B_T = \frac{w}{\sum_i p_i^2},
\]

where \(w\) is the bin width and \(p_i\) is the proportion of observations in
bin \(i\).

```{r tbreadth}
Tbreadth <- calc_Tbreadth(
  fish,
  bin_size = 0.1,
  print_results = FALSE
)
Tbreadth
```

The value is expressed in degrees Celsius. It becomes larger when the animal
uses a broader range of temperatures more evenly. It is a measure of selected
or experienced temperature breadth, not physiological thermal tolerance.
Always use the same `bin_size` when comparing animals.

## Plot the temperature distribution

`plot_coreT_histogram()` displays the percentage of observations within each
temperature bin. The dashed line marks the midpoint of the most frequently
occupied bin, and the subtitle reports thermal breadth calculated from the same
bins.

```{r coreT-histogram}
plot_coreT_histogram(fish, bin_size = 0.1)
```

The plot is returned invisibly, so it can also be saved or altered:

```{r save-plot, eval=FALSE}
p <- plot_coreT_histogram(fish)

ggplot2::ggsave(
  "core_temperature_histogram.png",
  plot = p,
  width = 7,
  height = 5
)
```

## Calculate behavioural and movement metrics

```{r behavioural-metrics}
shuttles <- calc_shuttles(fish, print_results = FALSE)
occupancy <- calc_occupancy(fish, print_results = FALSE)
distance <- calc_tot_distance(fish, print_results = FALSE)
tracking <- calc_track_accuracy(fish, print_results = FALSE)

results <- data.frame(
  metric = c(
    "Preferred temperature (°C)",
    "Lower avoidance temperature (°C)",
    "Upper avoidance temperature (°C)",
    "Effective thermal breadth (°C)",
    "Number of shuttles",
    "Seconds in DECR chamber",
    "Seconds in INCR chamber",
    "Total distance (cm)",
    "Proportion tracked"
  ),
  value = c(
    Tpref,
    Tavoid[1],
    Tavoid[2],
    Tbreadth,
    shuttles,
    occupancy[1],
    occupancy[2],
    distance,
    tracking
  )
)

knitr::kable(results, digits = 3)
```

Useful single-trial plots include:

```{r other-plots, eval=FALSE}
plot_T_segmented(fish)
plot_T_gradient(fish)
plot_tracking(fish)
plot_heatmap(fish)
plot_distance(fish)
plot_speed_coreT(fish)
```

## Exclude time from the start or end

Most calculation and plotting functions have matching
`exclude_start_minutes` and `exclude_end_minutes` arguments. For example, to
remove the first and final 10 minutes:

```{r time-exclusions}
calc_Tbreadth(
  fish,
  exclude_start_minutes = 10,
  exclude_end_minutes = 10,
  print_results = FALSE
)
```

Use the same exclusions for every metric that you want to compare.

## Analyse a folder of files

Import every ShuttleSoft `.txt` and `.csv` file in a folder:

```{r project-import, eval=FALSE}
all_fish <- read_shuttlesoft_project(
  directory = choose.dir()
)
```

Then calculate a row of results for each trial:

```{r project-results, eval=FALSE}
project_results <- calc_project_results(
  all_fish,
  Tpref_method = "median",
  Tavoid_percentiles = c(0.05, 0.95),
  Tbreadth_bin_size = 0.1
)
```

A metadata table can still be supplied when different files have different
trial start times. It only needs the columns relevant to your analysis:

```{r metadata-example, eval=FALSE}
metadata <- data.frame(
  file_name = c("Fish_1.txt", "Fish_2.txt"),
  trial_start = c("10:30:00", "14:15:00")
)

all_fish <- read_shuttlesoft_project(
  metadata = metadata,
  directory = choose.dir()
)
```


## Explore an existing project database

A project database contains one summary row per trial or individual. The
package includes an example database with 85 trials:

```{r import-project-database}
project_file <- system.file(
  "extdata",
  "project_database_example.csv",
  package = "ShuttleboxR"
)

project_data <- read_project_database(project_file)

dim(project_data)
head(project_data[c(
  "fileID", "mass", "Tpref", "Tavoid_lower", "Tavoid_upper",
  "Tpref_range", "tot_distance", "nr_shuttles"
)])
```

`read_project_database()` recognises names used by older ShuttleboxR versions.
For example, it automatically changes `study_ID` to `fileID`, `distance` to
`tot_distance`, and `shuttles` to `nr_shuttles`. This allows older project
results to work with the current analysis functions.

### Inspect distributions and possible outliers

`plot_histograms()` displays six project-level distributions and returns a table
of possible univariate outliers:

```{r project-histograms, fig.height=9, fig.width=8}
project_outliers <- plot_histograms(
  project_data,
  id_col = "fileID"
)

head(project_outliers)
```

### Examine relationships among metrics

```{r project-distance-shuttles, fig.height=5, fig.width=7}
plot_distance_vs_shuttles(
  project_data,
  label_points = FALSE
)
```

Related plots are available with:

```{r other-project-plots, eval=FALSE}
plot_limits_vs_distance(project_data, label_points = FALSE)
plot_limits_vs_shuttles(project_data, label_points = FALSE)
```

A correlation matrix can be produced for any selected numeric metrics:

```{r project-correlations, fig.height=7, fig.width=7}
correlation_values <- correlation_matrix(
  project_data,
  columns = c(
    "Tpref", "Tavoid_lower", "Tavoid_upper", "Tpref_range",
    "tot_distance", "nr_shuttles"
  )
)

round(correlation_values, 2)
```

### Principal component analysis

For PCA, first select the variables that are scientifically relevant to the
question. Including every numeric column automatically is rarely desirable.

```{r project-pca, fig.height=6, fig.width=7}
pca_data <- project_data[c(
  "fileID", "mass", "total_length", "Tpref", "Tavoid_lower",
  "Tavoid_upper", "Tpref_range", "tot_distance", "nr_shuttles"
)]

project_pca <- pca(
  pca_data,
  print_labels = FALSE,
  var_col = "Tpref"
)

project_pca$plots$biplot
```

The other PCA outputs are available as:

```{r other-pca-outputs, eval=FALSE}
project_pca$plots$screeplot
project_pca$plots$pc1contributionplot
project_pca$plots$varplot
project_pca$pca_scores
project_pca$pca_loadings
project_pca$outliers
```

The bundled example database was created before `Tbreadth` was added to
ShuttleboxR, so it does not contain that column. Thermal breadth cannot be
reconstructed from summary values alone. To add it to an existing project,
re-import the raw trial files and run `calc_project_results()` with a consistent
`Tbreadth_bin_size`.

## Recalculating core temperature

Most users should retain the `core_T` column already present in the ShuttleSoft
file. `calc_coreT()` is only needed when you deliberately want to apply a
separate thermal-lag model.

```{r recalculate-coreT, eval=FALSE}
fish <- calc_coreT(
  fish,
  mass = 12.4,
  a_value = 0.05,
  b_value = -0.25
)
```

The `a_value` and `b_value` coefficients cannot be estimated from the
ShuttleSoft file. They must come from an appropriate calibration or published
source.

## Common problems

**The package says that `trial_phase` is missing.** Import the file with
`read_shuttlesoft()` or run `file_prepare()` before requesting acclimation
exclusion.

**Everything is labelled as trial data.** This is expected when no
`trial_start` is supplied. Re-import the file with the actual trial start time
when acclimation needs to be removed.

**Thermal-breadth values change when `bin_size` changes.** This is expected
because the index is based on temperature bins. Choose a biologically and
instrumentally sensible bin width and keep it constant across animals.

**The package asks for `a_value` and `b_value`.** These are only required for
optional recalculation with `calc_coreT()`. They are not needed to analyse the
existing `core_T` values.
