---
title: "Getting started with ShuttleboxR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ShuttleboxR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

## Purpose and scope

ShuttleboxR bridges the gap between data acquisition and statistical analysis
in shuttle-box temperature experiments. Its purpose is to make the large files
produced by these experiments easier to organise, calculate consistently and
inspect for possible technical or behavioural problems.

The package follows three consecutive steps:

1. **Import and organise** the data.
2. **Calculate shuttle-box metrics** using explicit, reproducible settings.
3. **Inspect and troubleshoot** the data before formal statistical analysis.

This workflow has two connected branches:

- **Branch 1: single trials** calculates and inspects the recording from one
  fish.
- **Branch 2: multiple trials** compiles one row of metrics per fish and
  examines the study as a whole.

The branches are deliberately connected. Project-level functions can flag a
fish with unusual values, but the reason for that pattern can only be understood
by returning to the original single-trial record and inspecting its temperature,
tracking and movement data.

```text
Raw shuttle-box recordings
        |
        +-- Branch 1: one fish
        |      import and organise
        |             -> calculate metrics
        |             -> inspect the trial
        |
        +-- Branch 2: the complete project
               import all trials
                      -> compile one row per fish
                      -> inspect distributions and multivariate patterns
                                      |
                                      v
                             flag fish for review
                                      |
                                      v
                         return to Branch 1 plots
```

ShuttleboxR supports data preparation, quality control and exploration. It does
not select the inferential model for a study, and it does not automatically
decide that an unusual fish should be excluded.

```{r load-package}
library(ShuttleboxR)
```

## How a shuttle-box trial generates the data

A temperature shuttle box contains a warm and a cold chamber. During the dynamic
trial phase, the temperatures increase while the fish occupies the warm chamber
and decrease while it occupies the cold chamber. By moving between chambers, the
fish regulates the temperatures it experiences.

A typical recording therefore contains several linked data streams:

- time and trial phase;
- warm- and cold-chamber temperatures;
- chamber occupancy and shuttle events;
- fish position and distance moved; and
- estimated core body temperature (`core_T`).

These streams should be considered together. A plausible preferred temperature
is not enough to establish that a trial was successful if, for example, tracking
failed, temperature control was interrupted or the fish remained motionless.

## Workflow at a glance

| Branch and stage | Main question | Representative functions |
|---|---|---|
| Single trial: import and organise | Is this recording correctly structured and where does the experimental trial begin? | `read_shuttlesoft()`, `file_prepare()`, `inspect()` |
| Single trial: calculate | What thermal and behavioural metrics describe this fish? | `calc_Tpref()`, `calc_Tavoid()`, `calc_Tbreadth()`, `calc_gravitation()`, `calc_shuttles()`, `calc_occupancy()` |
| Single trial: inspect | Did the apparatus, tracking and fish behaviour produce an interpretable trial? | `plot_T_gradient()`, `plot_tracking()`, `plot_T_segmented()`, `plot_coreT_histogram()`, `plot_heatmap()` |
| Project: import and calculate | What is the common metric table across all fish? | `read_shuttlesoft_project()`, `calc_project_results()`, `read_project_database()` |
| Project: inspect | Which fish or trials differ from the rest of the study and require individual review? | `plot_histograms()`, project scatter plots, `correlation_matrix()`, `pca()` |

# Branch 1: work with one fish

## Step 1: import and organise a single trial

The simplest way to import a ShuttleSoft `.txt` file or `.csv` export is:

```{r choose-file, eval=FALSE}
fish <- read_shuttlesoft(file.choose())
```

`file.choose()` opens an ordinary file-selection window. By default,
`read_shuttlesoft()` also runs `file_prepare()`, which:

- creates elapsed time in seconds and hours;
- labels acclimation and trial phases;
- corrects dates when a trial passes midnight;
- identifies shuttle events; and
- converts the columns needed by later functions to the expected formats.

The current version no longer requires a separate metadata file for a single
trial. Values such as `trial_start` can be supplied directly, and the `core_T`
values already present in the ShuttleSoft file are retained unless the user
explicitly chooses to recalculate them.

This vignette uses a shortened recording included with the package:

```{r import-example}
example_file <- system.file(
  "extdata",
  "Fish_8_13_3_example.csv",
  package = "ShuttleboxR"
)

fish <- read_shuttlesoft(example_file)
```

Check the structure with:

```{r inspect-file}
dim(fish)
head(names(fish), 12)
inspect(fish)
```

### Marking acclimation and trial phases

Many experiments begin with a static acclimation phase followed by a dynamic
trial phase. Supply the clock time at which the dynamic trial began:

```{r trial-start, eval=FALSE}
fish <- read_shuttlesoft(
  file.choose(),
  trial_start = "13:30:00"
)
```

Rows before that time are labelled `"acclimation"`. When `trial_start` is
omitted, the entire recording is treated as trial data.

Most calculation functions can then use the same exclusion rule:

```{r exclude-acclimation, eval=FALSE}
calc_Tpref(fish, exclude_acclimation = TRUE)
calc_Tbreadth(fish, exclude_acclimation = TRUE)
plot_coreT_histogram(fish, exclude_acclimation = TRUE)
```

## Step 2: calculate metrics for the fish

### Primary data streams and summary metrics

Some calculations create data that other functions depend on. The most
important are:

- `core_T`, which is needed for thermal preference, avoidance, thermal breadth,
  gravitation time and time near limits; and
- cumulative `distance`, which is needed for total distance and movement speed.

ShuttleSoft normally supplies both. Use `calc_coreT()` or `calc_distance()` only
when the existing values need to be replaced or reconstructed.

The main summary metrics are:

```{r metric-table, echo=FALSE}
metric_table <- data.frame(
  Output = c(
    "Tpref",
    "Tavoid_lower / Tavoid_upper",
    "Tpref_range",
    "Tbreadth",
    "grav_time",
    "nr_shuttles",
    "tot_distance",
    "seconds_in_DECR / seconds_in_INCR",
    "t_near_limits",
    "core_T_variance",
    "track_accuracy"
  ),
  Interpretation = c(
    "Central selected temperature; median core_T by default",
    "Lower and upper percentiles of core_T",
    "Difference between upper and lower avoidance temperatures",
    "Effective width of the complete core_T frequency distribution",
    "Estimated time taken to approach the preferred temperature",
    "Number of moves between chambers",
    "Total distance moved during the selected period",
    "Time spent in the cold and warm chambers",
    "Proportion of observations near the system temperature limits",
    "Variation in core_T during the selected period",
    "Proportion of observations with successful position tracking"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(metric_table)
```

### Preferred and avoidance temperatures

`calc_Tpref()` summarises the selected temperatures. The default is median
`core_T`, although the mean and mode are available:

```{r tpref}
Tpref <- calc_Tpref(
  fish,
  method = "median",
  print_results = FALSE
)
Tpref
```

```{r tpref-methods}
calc_Tpref(fish, method = "mean", print_results = FALSE)
calc_Tpref(fish, method = "mode", print_results = FALSE)
```

`calc_Tavoid()` returns lower and upper percentiles of the `core_T`
distribution. The defaults are the 5th and 95th percentiles:

```{r tavoid}
Tavoid <- calc_Tavoid(
  fish,
  percentiles = c(0.05, 0.95),
  print_results = FALSE
)
Tavoid
```

Their difference is the central percentile range:

```{r tavoid-range}
Tpref_range <- Tavoid[2] - Tavoid[1]
Tpref_range
```

### Effective selected thermal breadth

`Tpref_range` uses only two points in the distribution. `calc_Tbreadth()` was
added later to provide a complementary measure that uses the frequency of all
experienced temperatures.

Temperatures are divided into equal-width bins and breadth is calculated as:

\[
B_T = \frac{w}{\sum_i p_i^2},
\]

where \(w\) is the temperature-bin width and \(p_i\) is the proportion of
observations in bin \(i\).

```{r tbreadth}
Tbreadth <- calc_Tbreadth(
  fish,
  bin_size = 0.1,
  print_results = FALSE
)
Tbreadth
```

A fish concentrated around a narrow set of temperatures has a small
`Tbreadth`. A fish that distributes its time broadly and relatively evenly has
a larger value. The result is expressed in degrees Celsius, but it is a measure
of selected or experienced thermal breadth, not physiological thermal
tolerance. Use the same `bin_size` for every fish being compared.

The percentile range and effective breadth should not be treated as
interchangeable. Two fish can have similar upper and lower avoidance
percentiles while differing substantially in how their time is distributed
between them.

### Behaviour, occupancy, movement and tracking

```{r behavioural-metrics}
shuttles <- calc_shuttles(fish, print_results = FALSE)
occupancy <- calc_occupancy(fish, print_results = FALSE)
distance <- calc_tot_distance(fish, print_results = FALSE)
tracking <- calc_track_accuracy(fish, print_results = FALSE)

results <- data.frame(
  metric = c(
    "Preferred temperature (°C)",
    "Lower avoidance temperature (°C)",
    "Upper avoidance temperature (°C)",
    "Preference percentile range (°C)",
    "Effective thermal breadth (°C)",
    "Number of shuttles",
    "Seconds in DECR chamber",
    "Seconds in INCR chamber",
    "Total distance (cm)",
    "Proportion tracked"
  ),
  value = c(
    Tpref,
    Tavoid[1],
    Tavoid[2],
    Tpref_range,
    Tbreadth,
    shuttles,
    occupancy[1],
    occupancy[2],
    distance,
    tracking
  )
)

knitr::kable(results, digits = 3)
```

Most metric functions also accept `exclude_start_minutes` and
`exclude_end_minutes`. Apply exclusions consistently when values will be
compared:

```{r time-exclusions}
calc_Tbreadth(
  fish,
  exclude_start_minutes = 10,
  exclude_end_minutes = 10,
  print_results = FALSE
)
```

## Step 3: inspect the single trial

The purpose of inspection is not simply to make attractive figures. It is to
ask whether the trial provides a valid and interpretable measure of the intended
behaviour.

### Did the apparatus and tracking system work?

Useful checks include:

```{r system-checks, eval=FALSE}
plot_T_gradient(fish)
plot_tracking(fish)
plot_T_segmented(fish)
```

These plots can reveal interruptions in heating or cooling, changes in the
temperature difference between chambers, tracking failures and an implausible
segmented estimate of gravitation time.

### Did the fish display interpretable behaviour?

The frequency distribution of `core_T` should be viewed alongside its summary
metrics:

```{r coreT-histogram}
plot_coreT_histogram(fish, bin_size = 0.1)
```

The dashed line marks the midpoint of the most frequently occupied temperature
bin, and the subtitle reports `Tbreadth` calculated using the same bins.

Position and activity plots can help distinguish active thermoregulation from
inactivity, tracking artefacts or prolonged occupancy of a particular area:

```{r behavioural-plots, eval=FALSE}
plot_heatmap(fish)
plot_distance(fish)
plot_interval(fish, column = "shuttle", interval_minutes = 30)
plot_speed_coreT(fish)
animate_movements(fish)
```

Time near the upper or lower system limits is especially important to examine.
It can indicate that the selected safety limits did not encompass the fish's
behaviour or that the fish was not responding to the gradient as expected.
Other metrics, such as shuttling frequency or total distance, are more strongly
species dependent and should be interpreted in an ecological context.

A plot can identify a reason for concern, but it cannot by itself establish that
a fish is invalid. Similar patterns across many fish may indicate a systematic
methodological issue or a genuine species-level behaviour.

# Branch 2: work with the complete project

For most studies, the final objective is not to describe one fish but to create
a consistent dataset across all trials. The project branch repeats the same
three stages at the level of the study.

## Step 1: import and organise multiple trials

Place the raw ShuttleSoft files in one directory and import them as a named
list:

```{r project-import, eval=FALSE}
all_fish <- read_shuttlesoft_project(
  directory = choose.dir()
)
```

When individual trials have different start times, provide an optional metadata
table:

```{r metadata-example, eval=FALSE}
metadata <- data.frame(
  file_name = c("Fish_1.txt", "Fish_2.txt"),
  trial_start = c("10:30:00", "14:15:00")
)

all_fish <- read_shuttlesoft_project(
  metadata = metadata,
  directory = choose.dir()
)
```

`compile_project_data()` combines all time-series rows into one long data frame.
This can be useful for plotting or checking the raw trajectories across several
trials. For one-row-per-fish results, use `calc_project_results()` instead.

## Step 2: create the project results database

```{r project-results, eval=FALSE}
project_results <- calc_project_results(
  all_fish,
  exclude_acclimation = TRUE,
  Tpref_method = "median",
  Tavoid_percentiles = c(0.05, 0.95),
  Tbreadth_bin_size = 0.1
)
```

The same settings are applied to every trial, while the acclimation boundary is
read separately from each file. This is important for making the resulting
metrics comparable.

An existing project-results CSV can be imported directly. The package includes
an example database containing 85 fish:

```{r import-project-database}
project_file <- system.file(
  "extdata",
  "project_database_example.csv",
  package = "ShuttleboxR"
)

project_data <- read_project_database(project_file)

dim(project_data)
head(project_data[c(
  "fileID", "mass", "Tpref", "Tavoid_lower", "Tavoid_upper",
  "Tpref_range", "tot_distance", "nr_shuttles"
)])
```

`read_project_database()` standardises several column names used by older
ShuttleboxR versions, such as `study_ID`, `distance` and `shuttles`.

The bundled project database predates `calc_Tbreadth()` and therefore has no
`Tbreadth` column. Thermal breadth cannot be reconstructed from summary values
alone; the raw temperature observations must be reprocessed with
`calc_project_results()`.

## Step 3: identify trials that require closer inspection

### Examine each metric across fish

`plot_histograms()` displays the distributions of key metrics and returns fish
that cross its univariate screening thresholds:

```{r project-histograms, fig.height=9, fig.width=8}
project_outliers <- plot_histograms(
  project_data,
  id_col = "fileID"
)

head(project_outliers)
```

An isolated value may reflect a technical problem, an unusual behavioural
strategy or genuine biological variation. The histogram is a screening tool,
not an exclusion rule.

### Examine relationships among metrics

Bivariate plots can reveal combinations of metrics that are more informative
than either value alone. For example, a fish with many shuttles but little total
distance may deserve a different interpretation from a fish with both high
shuttling and high movement.

```{r project-distance-shuttles, fig.height=5, fig.width=7}
plot_distance_vs_shuttles(
  project_data,
  label_points = FALSE
)
```

```{r other-project-plots, eval=FALSE}
plot_limits_vs_distance(project_data, label_points = FALSE)
plot_limits_vs_shuttles(project_data, label_points = FALSE)
```

A correlation matrix provides a broader summary of relationships:

```{r project-correlations, fig.height=7, fig.width=7}
correlation_values <- correlation_matrix(
  project_data,
  columns = c(
    "Tpref", "Tavoid_lower", "Tavoid_upper", "Tpref_range",
    "tot_distance", "nr_shuttles", "t_near_limits"
  )
)

round(correlation_values, 2)
```

### Use PCA as a multivariate screening tool

PCA summarises correlated project metrics and can flag fish that differ across
several variables simultaneously. Select variables that are scientifically
relevant rather than automatically including every numeric column:

```{r project-pca, fig.height=6, fig.width=7}
pca_data <- project_data[c(
  "fileID", "mass", "total_length", "Tpref", "Tavoid_lower",
  "Tavoid_upper", "Tpref_range", "tot_distance", "nr_shuttles"
)]

project_pca <- pca(
  pca_data,
  print_labels = FALSE,
  var_col = "Tpref"
)

project_pca$plots$biplot
```

The full output includes:

```{r other-pca-outputs, eval=FALSE}
project_pca$plots$screeplot
project_pca$plots$pc1contributionplot
project_pca$plots$varplot
project_pca$pca_scores
project_pca$pca_loadings
project_pca$outliers
```

The outlier table combines Mahalanobis-distance and DBSCAN screens. These
methods define unusual multivariate positions in different ways, but neither
establishes that a trial should be discarded.

## The recommended quality-control loop

The most important project-level step is to return flagged fish to the
single-trial branch:

```{r review-loop, eval=FALSE}
# Example identifier flagged by a histogram or PCA
fish_to_check <- all_fish[["Fish_17.txt"]]

# Was temperature control stable?
plot_T_gradient(fish_to_check)

# Was tracking adequate?
plot_tracking(fish_to_check)

# What did the thermal distribution look like?
plot_coreT_histogram(fish_to_check)

# Where did the fish spend its time?
plot_heatmap(fish_to_check)

# Inspect the calculated values together
calc_Tpref(fish_to_check, print_results = FALSE)
calc_Tavoid(fish_to_check, print_results = FALSE)
calc_Tbreadth(fish_to_check, print_results = FALSE)
calc_shuttles(fish_to_check, print_results = FALSE)
calc_extremes(fish_to_check, print_results = FALSE)
```

A transparent decision process is:

1. **Flag** an unusual pattern using project-level plots or PCA.
2. **Inspect** the original trial for technical, tracking and behavioural
   explanations.
3. **Contextualise** the pattern using the species' ecology and the wider
   dataset.
4. **Document** the rationale for retaining, qualifying or excluding the trial.

This helps avoid both retaining technically invalid trials and removing valid
biological variation simply because it is unusual.

## Optional recalculation of core temperature

Most users should retain the `core_T` column already present in ShuttleSoft
files. `calc_coreT()` is only needed when those values must be replaced using a
separate thermal-lag model:

```{r recalculate-coreT, eval=FALSE}
fish <- calc_coreT(
  fish,
  mass = 12.4,
  a_value = 0.05,
  b_value = -0.25
)
```

The `a_value` and `b_value` coefficients cannot be estimated from the
ShuttleSoft file. They must come from an appropriate calibration or published
source.

## Common problems

**The package says that `trial_phase` is missing.** Import the file with
`read_shuttlesoft()` or run `file_prepare()` before requesting acclimation
exclusion.

**Everything is labelled as trial data.** This is expected when no
`trial_start` is supplied. Re-import the file with the actual trial start time
when acclimation needs to be removed.

**Thermal-breadth values change when `bin_size` changes.** This is expected
because the index is based on temperature bins. Choose a biologically and
instrumentally sensible bin width and keep it constant across fish.

**The package asks for `a_value` and `b_value`.** These are only required for
optional recalculation with `calc_coreT()`. They are not needed to analyse the
existing `core_T` values.

**A fish is flagged as an outlier.** Treat the result as a prompt for
single-trial inspection, not as an automatic exclusion decision.
