---
title: "Getting started with ShuttleboxR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ShuttleboxR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  warning = FALSE,
  message = FALSE
)
```

## What ShuttleboxR is for

ShuttleboxR helps researchers move from raw shuttle-box recordings to a checked,
interpretable project dataset. It follows three steps:

1. **Import and organise** the raw data.
2. **Calculate shuttle-box metrics** consistently.
3. **Inspect and troubleshoot** the data before formal statistical analysis.

The same logic operates at two linked levels:

- **Branch 1: a single trial**, used to understand one fish in detail.
- **Branch 2: the complete project**, used to compare all fish and identify
  trials that deserve closer inspection.

```text
Raw shuttle-box recordings
        |
        +-- Branch 1: one fish
        |      import -> calculate -> inspect
        |
        +-- Branch 2: all fish
               compile metrics -> screen the project
                                      |
                                      v
                            identify fish to review
                                      |
                                      v
                         return to Branch 1 plots
```

The package does **not** automatically decide that an unusual fish should be
excluded. A project-level flag is the beginning of an investigation, not the
end of one.

```{r load-package}
library(ShuttleboxR)
```

## Choose your starting point

The correct import function depends on the structure of the data you already
have. ShuttleboxR has three entry points:

| Starting data | Import function | Resulting object | What happens next |
|---|---|---|---|
| One raw ShuttleSoft trial for one fish | `read_shuttlesoft()` | `fish`: time-by-time observations for one trial | Calculate and inspect that fish |
| A folder of raw ShuttleSoft trials, one file per fish | `read_shuttlesoft_project()` | `all_fish`: a named list of raw trials | Use `calc_project_results()` to create `project_data` |
| One existing project-summary CSV, with one row per fish | `read_project_database()` | `project_data`: the project summary table | Go directly to project-level plots and screening |

### Starting point 1: one raw trial file

```{r starting-point-single, eval=FALSE}
fish <- read_shuttlesoft(file.choose())
```

Use this route when the file contains the time-by-time ShuttleSoft recording for
one fish. The individual-level functions then calculate and inspect that trial.

### Starting point 2: a folder of raw trial files

```{r starting-point-folder, eval=FALSE}
all_fish <- read_shuttlesoft_project(
  directory = choose.dir()
)

project_data <- calc_project_results(all_fish)
```

The Windows dialogue created by `choose.dir()` displays folders rather than the
individual `.txt` or `.csv` files inside them. Navigate to the folder containing
all raw trial files and click **Select Folder**. `all_fish` is a list containing
the raw time series for each fish; `project_data` is the one-row-per-fish summary
table calculated from those trials.

### Starting point 3: an existing project-summary CSV

```{r starting-point-summary, eval=FALSE}
project_data <- read_project_database(file.choose())
```

Use this route when the CSV already contains one row per fish and columns such
as `Tpref`, `nr_shuttles`, `tot_distance` and `t_near_limits`. This route goes
straight to the project-level inspection functions. Do not import a summary
table with `read_shuttlesoft()` or `read_shuttlesoft_project()`, because those
functions expect raw time-series recordings.

```text
One raw fish file     -> read_shuttlesoft()         -> fish
Folder of raw files   -> read_shuttlesoft_project() -> all_fish
                                                |
                                                v
                                     calc_project_results()
                                                |
                                                v
                                           project_data

Existing summary CSV -> read_project_database()    -> project_data
```

An existing summary file can only provide metrics already stored in that file.
For example, `Tbreadth` cannot be reconstructed from `Tpref` and avoidance
temperatures alone; it must either be present in the summary table or be
recalculated from the raw trial recordings.

## The inspection logic

A fish may stand out for several different reasons, and those reasons lead to
different follow-up checks.

| Project-level signal | Possible interpretation | Useful single-trial checks |
|---|---|---|
| Very low distance and very few shuttles | Inactivity, poor health, stress, tracking failure, or genuinely sedentary behaviour | `plot_heatmap()`, `plot_tracking()`, `plot_distance()` |
| Many shuttles but little total distance | Repeated doorway crossings, localised movement, or a tracking artefact | `plot_heatmap()`, `animate_movements()`, `plot_tracking()` |
| High distance and high shuttling | High activity, active thermoregulation, or possible agitation/stress | `plot_interval()`, `plot_speed_coreT()`, `plot_heatmap()` |
| Substantial time near the upper limit | Preferred temperatures may exceed the programmed range, or the fish may not be regulating effectively | `plot_T_gradient()`, `plot_T_segmented()`, `plot_coreT_histogram()` |
| Substantial time near the lower limit | Preferred temperatures may fall below the programmed range, or the fish may not be regulating effectively | `plot_T_gradient()`, `plot_T_segmented()`, `plot_coreT_histogram()` |
| Multivariate PCA outlier | An unusual combination of several otherwise plausible metrics | Use PCA arrows and `outlier_details` to choose the most relevant single-trial plots |

The same pattern can have different meanings in different species. A low number
of shuttles may be normal for a sedentary species, while any sustained exposure
to a programmed safety limit is usually more concerning. Exclusion should be
based on a documented technical or biological reason, not simply on distance
from the project mean.

# Branch 1: analyse and inspect one fish

## Step 1: import and organise the recording

Select a ShuttleSoft `.txt` file or `.csv` export:

```{r choose-file, eval=FALSE}
fish <- read_shuttlesoft(file.choose())
```

`read_shuttlesoft()` imports and prepares the recording. It creates elapsed
seconds and hours, identifies shuttle events, converts required columns to the
expected formats, and labels acclimation and trial phases when a trial start is
provided.

```{r trial-start, eval=FALSE}
fish <- read_shuttlesoft(
  file.choose(),
  trial_start = "13:30:00"
)
```

When `trial_start` is omitted, the complete recording is treated as trial data.
The `core_T` values already present in the ShuttleSoft file are retained unless
the user explicitly recalculates them.

This vignette uses a clean, complete example recording included with the package:

```{r import-example}
example_file <- system.file(
  "extdata",
  "Fish_14_13_2.txt",
  package = "ShuttleboxR"
)

fish <- read_shuttlesoft(example_file)
inspect(fish)
```

## Step 2: calculate the main metrics

```{r basic-metrics}
Tpref <- calc_Tpref(fish, print_results = FALSE)
Tavoid <- calc_Tavoid(fish, print_results = FALSE)
Tpref_range <- Tavoid[2] - Tavoid[1]
Tbreadth <- calc_Tbreadth(fish, print_results = FALSE)
shuttles <- calc_shuttles(fish, print_results = FALSE)
distance <- calc_tot_distance(fish, print_results = FALSE)
tracking <- calc_track_accuracy(fish, print_results = FALSE)

single_results <- data.frame(
  metric = c(
    "Tpref (°C)",
    "Lower Tavoid (°C)",
    "Upper Tavoid (°C)",
    "Tpref range (°C)",
    "Tbreadth (°C)",
    "Number of shuttles",
    "Total distance (cm)",
    "Proportion successfully tracked"
  ),
  value = c(
    Tpref, Tavoid[1], Tavoid[2], Tpref_range,
    Tbreadth, shuttles, distance, tracking
  )
)

knitr::kable(single_results, digits = 3)
```

### Tpref, avoidance temperatures and Tpref range

`Tpref` is the central selected temperature. The default calculation is the
median `core_T`, although the mean and mode are available. `Tavoid_lower` and
`Tavoid_upper` are the lower and upper percentiles of the `core_T`
distribution; the defaults are the 5th and 95th percentiles. Their difference,
`Tpref_range`, is therefore the width of the central 90% of observations.

These metrics describe the centre and percentile boundaries of the distribution,
but they do not describe how observations are distributed *within* those
boundaries.

## What does Tbreadth mean?

`Tpref`, `Tavoid` and `Tbreadth` describe different parts of the temperature
distribution:

- `Tpref` describes **where** the distribution is centred.
- `Tavoid_lower` and `Tavoid_upper` describe percentile boundaries.
- `Tbreadth` describes **how far apart** the temperatures experienced by the
  fish were.

### Explain it like I am five

Imagine choosing two moments from the trial and looking at the fish's core
temperature at those moments.

- If the two temperatures are almost the same, their difference is small.
- If the two temperatures are far apart, their difference is large.

Now repeat this for every possible pair of moments and take the average of all
those differences. That average is `Tbreadth`.

A fish that stays at almost the same temperature will therefore have a
Tbreadth close to zero. A fish that regularly experiences temperatures far
apart will have a larger Tbreadth.

### The equation

For core-temperature observations \(T_1, T_2, \ldots, T_n\):

\[
Tbreadth = \frac{1}{n^2}
\sum_{i=1}^{n}\sum_{j=1}^{n}|T_i-T_j|
\]

In words, the function calculates the absolute temperature difference for all
pairs of observations and then takes their mean. This is also known as the
**Gini mean difference**. The implementation uses an efficient equivalent
calculation, so it does not need to construct an enormous table of every pair.

### Worked examples

| Temperature use | Tbreadth | Interpretation |
|---|---:|---|
| All observations at 20 °C | 0 °C | Every pair has the same temperature |
| 50% at 20 °C and 50% at 21 °C | 0.5 °C | Half the pairs differ by 1 °C and half by 0 °C |
| 50% at 10 °C and 50% at 20 °C | 5 °C | Half the pairs differ by 10 °C and half by 0 °C |
| 90% at 20 °C and 10% at 30 °C | 1.8 °C | The rare 30 °C observations increase breadth, but most pairs remain near 20 °C |

The 10 °C/20 °C example is important. Its Tbreadth is much larger than that of
a fish split between 20 °C and 21 °C because the two occupied parts of the
distribution are much farther apart. Unlike the previous bin-based definition,
the calculation therefore recognises the distance between peaks.

### What Tbreadth does not tell you

`Tbreadth` is **not centred on Tpref**. A value of 2 °C does not mean
`Tpref - 1 °C` to `Tpref + 1 °C`, and it does not define any particular lower
or upper temperature boundary.

It also cannot fully describe histogram shape. Two fish can have the same
Tbreadth even if one has a single broad peak and the other has two separate
peaks. Use the value together with `plot_coreT_histogram()`:

- `Tbreadth` summarises the overall spread in one number;
- the histogram shows whether that spread is narrow, broad, skewed, or
  multimodal.

`Tbreadth` is not a measure of physiological thermal tolerance.

```{r tbreadth-code}
calc_Tbreadth(
  fish,
  print_results = FALSE
)
```

## Step 3: inspect whether the trial is interpretable

Inspection should answer two separate questions:

1. Did the apparatus and tracking system work as intended?
2. Did the fish show behaviour that can reasonably be interpreted as
   thermoregulation?

### Chamber temperatures and system operation

```{r plot-gradient, fig.height=4.4}
plot_T_gradient(fish)
```

Look for a stable difference between the warm and cold chambers, plausible
heating and cooling rates, and no abrupt interruptions or impossible values. A
power cut or sensor failure may create a discontinuity that is obvious here but
not obvious from a single summary metric.

### Temperature trajectory, preference and gravitation

`plot_T_segmented()` displays `core_T` over time together with Tpref, avoidance
temperatures and the estimated gravitation breakpoint.

```{r plot-segmented, fig.height=5.2}
segmented_plot <- tryCatch({
  invisible(capture.output(
    p_segmented <- plot_T_segmented(
      fish,
      exclude_acclimation = FALSE,
      overlay_chamber_temp = TRUE
    )
  ))
  p_segmented
}, error = function(e) {
  ggplot2::ggplot(fish, ggplot2::aes(x = time_h, y = core_T)) +
    ggplot2::geom_line() +
    ggplot2::labs(
      title = "Core body temperature through the example recording",
      subtitle = paste("Segmented fit was not estimable:", conditionMessage(e)),
      x = "Time (h)", y = "Core body temperature (°C)"
    ) +
    ggplot2::theme_classic()
})
segmented_plot
```

A useful segmented plot should show a biologically plausible approach toward a
stable temperature region. Warning signs include an implausible breakpoint,
continued directional drift after the estimated gravitation time, extended
contact with system limits, or chamber temperatures that do not respond as
expected.

### Frequency distribution and thermal breadth

```{r core-temperature-histogram, fig.height=4.6}
plot_coreT_histogram(fish, bin_size = 0.1)
```

This plot is the visual partner to Tbreadth. Tbreadth increases when commonly
experienced temperatures are farther apart, while the histogram reveals *why*.
A narrow peak should produce a small Tbreadth. A broad peak, long tails, or
well-separated peaks can increase Tbreadth. The same Tbreadth can nevertheless
arise from different shapes, so inspect whether the distribution is symmetrical,
skewed, or multimodal. Multiple peaks may represent distinct phases of behaviour
and should be compared with `plot_T_segmented()` and the full temperature-through-
time record.

The `bin_size` argument here changes only how the histogram is drawn. It no
longer changes the value calculated by `calc_Tbreadth()`.

### Tracking and spatial behaviour

```{r tracking-plot, fig.height=4.4}
invisible(capture.output(
  tracking_plot <- plot_tracking(fish, interval_minutes = 20)
))
tracking_plot
```

```{r heatmap-plot, fig.height=5.2}
plot_heatmap(fish)
```

Poor tracking can inflate or suppress distance and may create false shuttles.
A heatmap concentrated in one tiny area can indicate inactivity, a tracking lock,
or a fish that did not engage with the thermal gradient. A broad heatmap is not
automatically better: location should be interpreted alongside chamber
occupancy, shuttling, temperature and species ecology.

# Branch 2: screen the complete project

## Step 1: create or import the project-level summary table

There are two routes into Branch 2.

### Route A: start from a folder of raw trial files

Import all raw `.txt` or `.csv` recordings from one folder:

```{r project-import, eval=FALSE}
all_fish <- read_shuttlesoft_project(
  directory = choose.dir()
)
```

The folder-selection dialogue shows folders rather than individual files. Select
the folder that contains all raw trials. The resulting `all_fish` object is a
named list containing one time-series dataset per fish.

Then calculate the same summary metrics for every fish:

```{r project-results, eval=FALSE}
project_data <- calc_project_results(
  all_fish,
  exclude_acclimation = TRUE,
  Tpref_method = "median",
  Tavoid_percentiles = c(0.05, 0.95)
)
```

The resulting `project_data` object contains one row per fish and can be passed
to all project-level plots and screening functions.

### Route B: start from an existing project-summary CSV

When a CSV already contains one row per fish and the required summary columns,
load it directly:

```{r existing-project-database, eval=FALSE}
project_data <- read_project_database(file.choose())
```

This route does not use `read_shuttlesoft_project()` or
`calc_project_results()`. It is appropriate only for a summary table, not a raw
ShuttleSoft recording.

The package includes an 85-fish example summary database, which is used for the
remainder of this vignette:

```{r import-project-database}
project_file <- system.file(
  "extdata",
  "project_database_example.csv",
  package = "ShuttleboxR"
)
project_data <- read_project_database(project_file)
```

The example database predates `calc_Tbreadth()`, so it does not contain that
metric. Tbreadth must be recalculated from the original time-series data; it
cannot be reconstructed from Tpref and avoidance temperatures alone.

## Step 2: begin with univariate distributions

```{r project-histograms, fig.height=9, fig.width=8}
univariate_flags <- plot_histograms(project_data)
```

The returned table identifies fish that are unusual for individual metrics.
This is useful for finding, for example, exceptionally low distance, an unusual
Tpref or substantial time near limits. A univariate flag does not reveal whether
the value is technically invalid or biologically meaningful.

## Step 3: inspect upper and lower limit exposure separately

The combined `t_near_limits` value can conceal whether a fish repeatedly reached
the upper limit, the lower limit, or both. The separate plot is therefore an
important quality-control step.

```{r upper-lower-extremes, fig.height=5.4}
extreme_review <- plot_upper_vs_lower_extremes(
  project_data,
  return_cases = TRUE
)
extreme_review$plot
```

```{r upper-lower-cases}
knitr::kable(extreme_review$cases, digits = 2)
```

By default, the dashed lines mark **10% of analysed observations** near each
limit. This direct percentage threshold is easier to interpret than a project-level
IQR rule and remains meaningful when most fish have zero exposure. A point is
highlighted only when it exceeds the relevant threshold; any small non-zero
value is no longer treated as "high" exposure.

Interpretation:

- **Above upper-limit threshold:** the fish may prefer warmer conditions than the
  programmed range permits, or it may have failed to move away from the warm
  extreme.
- **Above lower-limit threshold:** the analogous concern at the cold end.
- **Above thresholds at both limits:** may indicate unusually wide exploration,
  unstable system behaviour, tracking problems, or settings poorly matched to
  the species.

The thresholds can be changed explicitly, for example:

```r
plot_upper_vs_lower_extremes(
  project_data,
  lower_limit_threshold = 5,
  upper_limit_threshold = 5
)
```

Project-relative alternatives remain available through
`limit_method = "quantile"` or `limit_method = "iqr"`, but these should be used
with care in datasets containing many zero values.

The next checks should be `plot_T_gradient()`, `plot_T_segmented()` and
`plot_coreT_histogram()` for the flagged fish.

## Step 4: use bivariate plots to distinguish different outlier types

A bivariate plot can identify three kinds of cases:

1. **Marginal outliers:** extreme on one axis, such as very low distance.
2. **Corner cases:** extreme on both axes, such as low distance and low
   shuttling.
3. **Unusual combinations:** values that are individually plausible but form an
   unexpected combination, such as many shuttles with little total movement.

The optional review guides below use the 5th and 95th project quantiles for
movement and shuttling. By default, limit exposure is flagged only when more
than **10% of analysed observations** occurred near the programmed limits. This
absolute threshold is a transparent starting point, not a universal biological
cutoff, and can be changed with `limit_threshold`.

The older project-relative approaches remain available with
`limit_method = "quantile"` or `limit_method = "iqr"`. The absolute method is
the default because IQR and quantile thresholds can collapse towards zero when
most fish never approach the limits.

### Distance versus shuttling

This is a useful first project-level comparison because total distance describes
overall activity, whereas the number of shuttles describes movement between the
two chambers. The plot can therefore distinguish inactivity from active movement
that does not involve thermal regulation.

```{r distance-shuttle-review, fig.height=5.8, fig.width=7.5}
distance_shuttle_review <- plot_distance_vs_shuttles(
  project_data,
  highlight_cases = TRUE,
  return_cases = TRUE
)
distance_shuttle_review$plot
```

```{r distance-shuttle-cases}
knitr::kable(distance_shuttle_review$cases, digits = 2)
```

Particularly useful patterns are:

- **Low movement + low shuttling:** inspect inactivity and tracking.
- **Low movement + high shuttling:** inspect the doorway region, false crossings
  and repeated local movement.
- **High movement + low shuttling:** the fish may be active within one chamber
  but not regulating through chamber changes.
- **High movement + high shuttling:** may represent highly active
  thermoregulation or agitation.

### Limit exposure versus distance

```{r limits-distance-review, fig.height=5.8}
limits_distance_review <- plot_limits_vs_distance(
  project_data,
  highlight_cases = TRUE,
  return_cases = TRUE
)
limits_distance_review$plot
```

A fish **above the limit-exposure threshold with low movement** may have
remained near a limit without responding. A fish **above the threshold with
high movement** may have been actively searching but unable to reach a suitable
temperature, which can indicate that the programmed range was inappropriate.

### Limit exposure versus shuttling

```{r limits-shuttles-review, fig.height=5.8}
limits_shuttles_review <- plot_limits_vs_shuttles(
  project_data,
  highlight_cases = TRUE,
  return_cases = TRUE
)
limits_shuttles_review$plot
```

Exposure above the threshold with few shuttles suggests little corrective
behaviour. Exposure above the threshold with many shuttles suggests that the
fish was responding but the available gradient or limits may not have allowed
it to stabilise.

## Step 5: view the wider pattern with a bivariate matrix

The targeted plots above are best for identifying and labelling particular fish.
A bivariate matrix provides a wider overview by displaying every pairwise
relationship among a selected set of project metrics in one figure.

```{r project-bivariate-matrix, fig.height=8, fig.width=8}
project_correlations <- correlation_matrix(
  project_data,
  columns = c(
    "Tpref",
    "Tpref_range",
    "grav_time",
    "tot_distance",
    "nr_shuttles",
    "t_near_limits"
  )
)
```

Read the matrix as follows:

- Variable names appear along the diagonal.
- The lower panels show the individual fish and a smooth trend for each pair of
  metrics.
- The upper panels show the Pearson correlation coefficient. Values near 1 or
  -1 indicate strong positive or negative relationships; values near 0 indicate
  little linear association.
- A point separated from the main cloud in one lower panel may represent an
  unusual combination of otherwise plausible measurements.

The matrix is particularly useful for detecting redundant variables before PCA.
For example, two strongly correlated metrics can contribute very similar
information. It does not label individual fish, so a suspicious relationship
should be followed up with the corresponding targeted bivariate plot and then
with the original single-trial record.

## Step 6: interpret PCA as a map of multivariate behaviour

PCA is useful when no single metric fully explains why a fish is unusual. It
compresses correlated metrics into principal components.

Use a deliberately chosen set of variables. Avoid automatically including every
numeric column, and avoid including several mathematically redundant variables
unless that redundancy is scientifically intended.

```{r project-pca, fig.height=5.8, fig.width=7.5}
project_pca <- pca(
  project_data,
  variables = c(
    "Tpref",
    "Tpref_range",
    "grav_time",
    "tot_distance",
    "nr_shuttles",
    "t_near_max",
    "t_near_min"
  ),
  mahalanobis_th = 0.99,
  dbscan_th = 1.5,
  dbscan_minPts = 4,
  flag_rule = "both",
  print_labels = FALSE,
  highlight_outliers = TRUE
)

project_pca$plots$screeplot
```

### Reading the scree plot

The scree plot shows the percentage of total project variation represented by
each principal component. PC1 is the largest single axis of variation, PC2 the
next largest, and so on. The PC1-PC2 biplot is only a two-dimensional view. If
PC1 and PC2 explain a modest fraction of total variation, fish can be unusual in
later components even when they do not appear extreme on the biplot.

```{r pca-biplot, fig.height=6.4, fig.width=8}
project_pca$plots$biplot
```

### Reading the biplot

- **Points close together** have similar combinations of metrics.
- **Points far apart** have different multivariate profiles.
- **An arrow points toward increasing values** of that metric.
- **Longer arrows** are represented more strongly in the displayed PC1-PC2
  plane.
- **Arrows pointing in similar directions** indicate positively associated
  metrics; opposing arrows indicate negative association; near-right angles
  indicate weak association in this plane.
- A fish lying far in the direction of an arrow is likely to have a relatively
  high value for that metric. A fish in the opposite direction is likely to
  have a relatively low value.

Flagged fish are circled and labelled. The arrows suggest what to inspect next,
but `outlier_details` makes that link more explicit:

```{r pca-outlier-details}
knitr::kable(project_pca$outlier_details, digits = 2)
```

`potential_drivers` lists the original measurements that are most unusually
high or low for each flagged fish, expressed in project standard deviations.
For example, a fish driven by high `t_near_max` should be checked for upper-limit
exposure, whereas one driven by low `tot_distance` should be checked for
inactivity or poor tracking.

### Mahalanobis distance and DBSCAN are complementary

- **Mahalanobis distance** asks whether a fish is far from the multivariate
  centre after accounting for correlations among metrics. It uses all retained
  PCA dimensions, not only the visible biplot.
- **DBSCAN** asks whether a fish lies in a locally sparse region of the PC1-PC2
  map. It can detect isolated points even when the project contains more than
  one cluster.

Agreement between the methods is a strong reason for closer inspection, but
neither method proves invalidity. Disagreement is also informative: a fish can
be globally unusual without being locally isolated, or locally isolated without
being far from the overall centre.

The thresholds and flagging rule control sensitivity:

- Increasing `mahalanobis_th` makes the Mahalanobis screen more conservative.
  For example, `0.99` requires a fish to be farther from the multivariate
  centre than `0.975`.
- Increasing `dbscan_th` allows points farther apart to remain in the same local
  neighbourhood, and therefore usually reduces the number classified as noise.
- `dbscan_minPts` sets how many neighbouring points are needed to form a dense
  region. Its effect depends on sample size and the structure of the data, so it
  should be checked rather than interpreted in isolation.
- `flag_rule = "both"` highlights only fish detected by both methods and is the
  most conservative choice. `"either"` highlights fish detected by at least one
  method, while `"mahalanobis"` and `"dbscan"` use only the named screen.

The method-level detections remain available in `project_pca$outliers`. The
fish-level table `project_pca$screening` shows which method detected each fish
and whether it was selected under the chosen `flag_rule`:

```{r pca-screening}
knitr::kable(
  subset(project_pca$screening, mahalanobis | dbscan),
  digits = 2
)
```

PCA results depend on the variables included. Adding or removing variables can
rotate the axes and substantially change the plot even when the fish have not
changed. For that reason, the variables are supplied explicitly above and
should always be reported with the thresholds and flagging rule.

## Step 7: return flagged fish to the raw trial

```{r review-loop, eval=FALSE}
fish_to_check <- all_fish[["Fish_17.txt"]]

plot_T_gradient(fish_to_check)
plot_tracking(fish_to_check)
plot_T_segmented(fish_to_check)
plot_coreT_histogram(fish_to_check)
plot_heatmap(fish_to_check)
plot_distance(fish_to_check)

calc_Tpref(fish_to_check, print_results = FALSE)
calc_Tavoid(fish_to_check, print_results = FALSE)
calc_Tbreadth(fish_to_check, print_results = FALSE)
calc_shuttles(fish_to_check, print_results = FALSE)
calc_extremes(fish_to_check, print_results = FALSE)
```

A defensible decision process is:

1. **Flag** an unusual metric or combination of metrics.
2. **Inspect** the original temperature, tracking and movement records.
3. **Identify a reason**, such as a power interruption, tracking failure,
   prolonged immobility, inappropriate limits, or behaviour inconsistent with
   the intended assay.
4. **Compare** the pattern with the rest of the project and the ecology of the
   species.
5. **Document** whether the fish is retained, excluded, or analysed in a
   sensitivity analysis.

A fish should not be excluded merely because it is statistically unusual.
Unusual values can be genuine biological variation. Conversely, a technically
flawed trial can require exclusion even when its summary metrics do not appear
extreme.

## Getting help

Open the help page for any function with, for example:

```{r help-pages, eval=FALSE}
?calc_Tbreadth
?plot_distance_vs_shuttles
?pca
```

List all package help pages with:

```{r package-help, eval=FALSE}
help(package = "ShuttleboxR")
```
